---
title: "Quiz 2 question 2 plots"
author: "Phil Allen"
date: "13 February 2016"
output: html_document
---


```{r, echo=FALSE,results='hide', message=FALSE}
library(caret)
library(AppliedPredictiveModeling)
library(Hmisc)
library(ggplot2)
data(concrete)
```

Plot the outcome compressive strength by row index, to see if the data has been accumulated from sets with different characteristics.

```{r, echo=FALSE}
plot(mixtures$CompressiveStrength)
```

There seem to be diagonal bands, with discontinuities around rows 200, 400, 550 and 900 in the data set 

```{r}
# Cut compressive strength into five even-size bins.
mixtures$CutCS <- cut2(mixtures$CompressiveStrength,g=5)
# Add index column to help qplot.
mixtures$idx <- 1:nrow(mixtures)
```

The variables of interest are: `r sort(names(mixtures)[1:8])`

Plot them against index, colouring by the binned compressive strength.

```{r, echo=FALSE}
qplot(idx, Age, data=mixtures, colour=CutCS)
```
```{r, echo=FALSE}
qplot(idx, BlastFurnaceSlag, data=mixtures, colour=CutCS)
```
```{r, echo=FALSE}
qplot(idx, Cement, data=mixtures, colour=CutCS)
```
```{r, echo=FALSE}
qplot(idx, CoarseAggregate, data=mixtures, colour=CutCS)
```
```{r, echo=FALSE}
qplot(idx, FineAggregate, data=mixtures, colour=CutCS)
```
```{r, echo=FALSE}
qplot(idx, FlyAsh, data=mixtures, colour=CutCS)
```
```{r, echo=FALSE}
qplot(idx, Superplasticizer, data=mixtures, colour=CutCS)
```
```{r, echo=FALSE}
qplot(idx, Water, data=mixtures, colour=CutCS)
```

None of these variables clearly aligns with the discontinuities in the outcome measure. This may suggest a missing predictor variable.


